A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-09 paper.tex
More updates

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell). The edges are ``annotated''
with the genome coordinates over which inheritance occurs.
Fig~\ref{fig-ancestry-resolution} is an instructive example
from the literature \citep{wiuf1999recombination} and illustrates
how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).

There are two distinct annotation schemes possible, and the distinction is subtle
and important. The first scheme is to use the annotations to describe the complete
set of genomic intervals directly inherited by the descendant node from the ancestor
node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
As we discuss in section XXX, this second scheme has many advantages
in terms of efficiency.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
two gARGs with alternative edge annotation schemes.
(A) A conventional Griffiths eARG. Square nodes represent events, with
each recombination (red) annotated with the position of a single breakpoint
along the 7-unit-long chromosome.
(B) A gARG with direct-inheritance edge annotations. Oval nodes represent
genomes and annotations reflect ``local'' transmission of all genetic material
between parent and child. Edges that do not span the full genome are visually
highlighted, showing how breakpoint positions can be encoded.
(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
the relative importance of different transmission pathways to the samples. Nodes
\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
to events which have no impact on the genetic ancestry of the samples.
}
\end{figure}

@@

Given the direct-inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}B),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the sample-resolved material for a node is the
union of the sample-resolved intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their direct-inheritance intervals
and the just-computed sample-resolved intervals for the node.
For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
joining \textsf{i} to its ancestor \textsf{k} then carries
material on $(0, 2]$ since although \textsf{i} directly inherits interval
$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
ancestral to the sample. For each node, we also keep track of the number of
descendant samples over each genomic interval, and stop tracking that
interval once the lineages from all samples have coalesced. For example,
the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
already fully coalesced (indicated by a fully black genomic region).
This process is directly analogous to Hudson's simulation algorithm
(Appendix XXX).

In practice, the backwards-in-time Griffiths process that produces
ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
dominated by nodes such as \textsf{g} and \textsf{j} which contain information
about inheritance in non-ancestral material, as well as nodes such as
\textsf{o} and \textsf{q} which track details of ancestry even once the
local genomic region has fully coalesced. Instead, simulating
ARGs from the coalescent with recombination is usually done by
running Hudson's algorithm to produce a sample-resolved ARG directly.
Moreover, when inferring ARGs from genetic data, there is no information
in the sample set that can be used to resolve these nodes, so that they
are essentially unknowable.

Nevertheless, there is a practical use for converting a direct-inheritance
gARG to a sample resolved version. It is the key algorithm underlying recent
advances in forwards-in-time simulation.
\cite{kelleher2018efficient} developed a method in which forwards-in-time
simulations are performed by recording all parent-child genetic relationships
(i.e.~annotating the edges in the simulated pedigree with the genomic
intervals inherited by a child from a parent) and periodically ``simplifying''
this structure to retain only ancestral relationships that are relevant
to the ancestry of the current population. In many situations, using this approach to
record the full genetic ancestry of the current population is orders of magnitude faster
than the alternative of tracking neutral genetic variation directly~\citep{kelleher2018efficient,haller2018tree}.

@@

>>>>

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell). The edges are ``annotated''
with the genome coordinates over which inheritance occurs.
There are two different ways in which we use these inheritance
annotations, and the distinction is subtle and important. The difference
is illustrated in Fig~\ref{fig-ancestry-resolution}.

% TODO majorly condense this caption, it's much too long.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
TODO DELETE A? Or keep it? Maybe this would help some hard-core eARGers?
The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
two gARGs with alternative edge annotation schemes.
% (A) A conventional Griffiths eARG. Square nodes represent events, with
% each recombination (red) annotated with the position of a single breakpoint
% along the 7-unit-long chromosome.
(A) A gARG with edges annotated by . Oval nodes represent
genomes and annotations reflect ``local'' transmission of all genetic material
between parent and child. Edges that do not span the full genome are visually
highlighted, showing how breakpoint positions can be encoded.
(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
the relative importance of different transmission pathways to the samples. Nodes
\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
to events which have no impact on the genetic ancestry of the samples.
}
\end{figure}

The simplest interpretation of the inheritance intervals $I$ for
an ancestor genome $v$ and descendant $u$, is that $I$ defines the
regions of genome that $u$ inherited from $v$. For example,
node $o$ is the recombinant descendant of nodes $p$ and $q$ in
Figure~\ref{fig-ancestry-resolution}A, which shows that it in inherited
everything to the left of position $1$ from $p$ and the rest of
its genome from node $q$.
Thus, these edge annotations are concerned only with the genetic relationship
between the nodein question, and are not affected by the rest of the graph.
We will thefore refer to inheritance annotations of this type as ``local inheritance
annotations''. Graphs of this type result from forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree} or the
``Big ARG'' stochastic process (Appendix XXX).

ARGs are most often thought of as a \emph{retrospective} structure however,
defined with respect to some set of (usually, contemporary) sample genomes.
In this case we are not explicitly interested in the
entire interval of genome descendant $u$ inherited from ancestor $v$,
but rather the subset of that interval that is \emph{ancestral to our
sample}.
Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
reduce the annotation intervals to those that are ancestral to the
sample (note that node $q$ is now ommitted because the MRCA of the sample
has already been found for the full genome at node $p$).
Tracking such
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
is usually much more useful and, indeed, all that is possible to
have information about when inferring ARGs from contemporary samples.
We refer to a gARG with inheritance annotations of this type as
``sample resolved''.

% Fig~\ref{fig-ancestry-resolution} is an instructive example
% from the literature \citep{wiuf1999recombination} and illustrates
% how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
% a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).

% There are two distinct annotation schemes possible, and the distinction is subtle
% and important. The first scheme is to use the annotations to describe the complete
% set of genomic intervals directly inherited by the descendant node from the ancestor
% node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
% describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
% of the sample that is propagated through
% the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
% As we discuss in section XXX, this second scheme has many advantages
% in terms of efficiency.

@@

Given a gARG with direct-inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}B),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the sample-resolved material for a node is the
union of the sample-resolved intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their direct-inheritance intervals
and the just-computed sample-resolved intervals for the node.
For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
joining \textsf{i} to its ancestor \textsf{k} then carries
material on $(0, 2]$ since although \textsf{i} directly inherits interval
$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
ancestral to the sample. For each node, we also keep track of the number of
descendant samples over each genomic interval, and stop tracking that
interval once the lineages from all samples have coalesced. For example,
the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
already fully coalesced (indicated by a fully black genomic region).

% In practice, the backwards-in-time Griffiths process that produces
% ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
% is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
% dominated by nodes such as \textsf{g} and \textsf{j} which contain information
% about inheritance in non-ancestral material, as well as nodes such as
% \textsf{o} and \textsf{q} which track details of ancestry even once the
% local genomic region has fully coalesced. Instead, simulating
% ARGs from the coalescent with recombination is usually done by
% running Hudson's algorithm to produce a sample-resolved ARG directly.
% Moreover, when inferring ARGs from genetic data, there is no information
% in the sample set that can be used to resolve these nodes, so that they
% are essentially unknowable.

The process of converting a gARG with local-inheritance annotations
to its sample-resolved equivalent is also known as ``simplification'',
and is the key algorithm underlying recent
advances in forwards-in-time
simulation~\citep{kelleher2018efficient,haller2018tree}.
It has many useful properties, as we discuss in the next section.

@@

% FIXME putting this here for now just to see if this will work.

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
% The final vital
% detail is that we associate a breakpoint with each recombination
% event.
% This information is sufficient to uniquely
% define the local genealogical trees at every position along the genome,
% which is the basic requirement for a data structure encoding
% genome-wide genetic ancestry.
As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure),
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
The order in which parents are listed at a recombination node is
therefore significant, telling us
from which parent the segments of genome to the left and right of the breakpoint
were inherited.

This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event older than a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic.

\begin{figure}
\centering
% \begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] at (0, 0) {};
\node (s1) [greynode] at (3, 0) {};
\node (s2) [greynode] at (6, 0) {};
\node (s3) [greynode] at (3, 1) {};
\node (s4) [greynode] at (1, 2) {};
\node (s5) [greynode] at (5, 3) {};
\node (s6) [greynode] at (3, 4) {};

% \node [anchor=north west] at (0,6) {A};
\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) -- (s3);
\draw (s0) |- (s4);
\draw (s4) -- (2,2) |- (s3);
\draw (s4) |- (s6);
\draw (s3) -- (4,1) |- (s5);
\draw (s2) |- (s5);
\draw (s5) |- (s6);

% \node [anchor=north west] at (9,6) {B};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

% \node [anchor=north west] at (9,6) {C};
% \node [nodelabel,anchor=north west] at ($(10,5)$) {
% \begin{tabular}{ll}
% Edges & Intervals\\
% \hline
% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A graphical depiction of a classical event ARG and a concrete
encoding of the information. Node \noderef{d} is a recombination
event as it is associated with a non-null breakpoint, and has
two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
common ancestor events.
}
\end{figure}

% What do we mean by an eARG, formally?
To be concrete, we can formally define the
classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(This is equivalent to having different ``types''
of node, where breakpoints are associated with recombination
nodes only).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
% For each recombination node $u$ we must
% have exactly two edges in $e$ such that $u$ is the child.
This encoding of an eARG is equivalent to the output
of programs such as
\texttt{ARGweaver}~\citep{rasmussen2014genome}
and \texttt{KwARG}~\citep{ignatieva2021kwarg},
and an example is given in Fig.~\ref{fig-arg-data-structure}A.
Note that the description above captures only the
graph topology: if we also wish to know the branch lengths we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
